### 2022-10-13 AISTATS2023/paper.tex
latest

We define UALE and provide an interval-based formulation that
partitions the feature domain into non-overlapping intervals
(bin-splitting) and defines the bin-effect and the
bin-uncertainty. Then, we present UALE approximation, formally proving
the bin-splitting requirements for an unbiased estimation of the
uncertainty. Based on this, we present an optimization method to
determine an appropriate variable-width partitioning.

@@

UALE defines the effect as a pair that consists of the average effect
and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
and the accumulated uncertainty, as defined in
Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates
an example \todo{Describe after putting the figure}.

\begin{figure}
  \centering
  \caption{UALE-concept-figure}
  \label{fig:UALE-figure}
\end{figure}

@@

In this simulation, we focuse on illustrating the advantages of UALE's
variable size bin-splitting against fixed-size bin-splitting.

\paragraph{Example set-up.}
The purpose of this experiment is two-fold. First, we show that UALE
automatically provides an approximation that is as accurate (or even
more accurate) than the best fixed-size approximation, among all
possible \(K\). Second, we illustrate that UALE favors the creation of
wide intervals when they are not adding cumulative bin-error
\(\mathcal{E}^2_{\mathcal{Z}}\). Wide intervals facilitate the
interpretation of the average effect and its uncertainty.

For measuring the accuracy of the approximation, we generate a very
big dataset with dense sampling (\(N=10^6\)) and we use dense
fixed-size bins (\(K=10^3\)). We treat this estimation as the
ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
and we compare the fixed-size estimation (for several \(K\)) against
UALE's. We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We, also, denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) the sequence obtained by
automatic bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the
fixed-size splitting with \(K\) bins. The evaluation is based on the
Mean Absolute Error (MAE) in estimating the bin effect \(\mu\) and the
uncertainty \(\sigma\) accross bins, i.e.,

\begin{equation}
  \label{eq:eval_met_1}
  \mathcal{L}^{\mu} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} | \mu(z_{k-1}, z_k) - \hat{\mu}(z_{k-1}, z_k) |
\end{equation}

\begin{equation}
  \label{eq:eval_met_2}
  \mathcal{L}^{\sigma} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
    \mathcal{Z}} | \sigma(z_{k-1}, z_k) - \hat{\sigma}(z_{k-1}, z_k) |
\end{equation}

We obtaint the ground truth bin-effect, \(\mu(z_{k-1}, z_k)\), and the
ground truth uncertainty, \(\sigma(z_{k-1}, z_k)\), using the average
of the dense fixed-size bins that are within in the interval
\([z_{k-1}, z_k]\). We also provide
\(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \) for facilitating the
interpretation of the uncertainty error, and we call it mean residual error.

@@

Here, the black-box function is \(f(\mathbf{x}) = a_1x_1 + x_1x_2\),
with \(5\) different-width regions, i.e., \(a_1\) equals to
\(\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the sequence
\(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). We are interested in the effect of
\(x_1\) which is \(f^{\mathcal{GT}}_{\mu}(x_1) = a_1x_1\) in the
regions defined above and the uncertainty is
\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\).

As we observe in the top left of Figure~\ref{fig:ex-synth-2-1}, UALE's
bin-splitting separates the fine-grained bins, e.g.~regions
\([0.4, 0.45]\), \([0.45, 0.5]\), and unites (most) constant-effect
regions into a single bin, e.g.~region \([0.5, 1]\). This explains the
fact that UALE achieves lower MAE \(\mathcal{L}^{\mu}\),
\(\mathcal{L}^{\sigma}\) compared to fixed-size bins for all \(K\).

@@

Here, the black-box function is
\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the effect is
\(f^{\mathcal{GT}}_{\mu}(x_1) = 4 x_1^2\) and the uncertainty
\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\). Due to the
non-linear effect, wide bins increase the mean residual error,
\(\mathcal{L}^{\rho}\) making the uncertainty approximation
biased. This can be clearly observed in in the bottom-right of
Figure~\ref{fig:ex-synth-2-2}. On the other hand, narrow bins lead to
worse approximation due to the limited number of
samples. Interestingly, UALE's bin splitting manages to compromise
these conflicting objectives and provide an accurate estimation, as we
see in the bottom-left of Figure~\ref{fig:ex-synth-2-2}.

@@

In real-world datasets, it is infeasible to obtain the ground truth FE
for evaluation. We observe in Figure \ref{fig:ex-real-1} that in broad
terms UALE and PDP-ICE plots aggree on the average effect and the
uncertainty. It is interesting to note that UALE has selected bins of
varying size as the optimal partiotioning. This demonstrates an
important advantage of UALE; through the bin splitting process we can
identify wide intervals of the feature domain with similar level of
uncertainty.

% We selected the particular experiment, because, in
% broad terms, UALE and PDP-ICE plots aggree in the estimation of the
% average effect and uncertainty. Therefore, we can focus on juding the
% quality of the information provided by the two methods.

% In Figure\ref{fig:ex-real-1}, we observe, from top to bottom, the
% effects for the latitude, population and the median income. The effect
% of UALE and PDP-ICE are similar for the population and the median
% income. The population has a negative impact that progressively
% decreases: from 400 to 1500 people the house value decreases with a
% rate of \(-150 (\pm 140)\) dollars per added person, a rate that
% decreases from \(-80 (\pm 80)\) to \(-60 (\pm 60)\) dollars per added
% person as we move from 1500 to 2800 people. The level of uncertainty
% indicates significant variance in absolute value of the rate, but in
% the grant majority of instances the rate is negative. With the same
% inspection, we observe that the median annual income has a positive
% impact on the value (all numbers are thousands of dollars):
% \(20\pm 15\) per \(10\) of added median income for incomes in
% \([8, 15]\), \(32 \pm 20\) per \(10\) added income in \([15, 60]\) and
% \(40 \pm 15\) per \(10\) added income in \([60, 70]\). The uncertainty
% indicates that there are less heterogeneous effects about the median
% income compared to the number of people. In both cases, we can end-up
% to the same conclusion by inspecting the PDP-ICE plots. For the
% latitude, there is a small difference in the explanations for the
% region \([32, 35]\), where UALE estimates a less negative slope with
% less uncertainty than PDP, while the explanations are similar for the
% range \([35,39]\), where both methods reveal an increase in the
% uncertainty around the feature value \(37.5\).

% In general, we observe that UALE complements ALE in the quantification
% of the heterogeneous effects, similarly to as ICE complements PDP. The
% automatic extraction of constant-effect and constant-uncertainty
% regions provided by UALE is helpful for an easier interpretation.

% On
% the other hand, ICE plots sometimes are more descritive due to
% illustrating the type of heterogeneity, whereas UALE quantifies only the
% level (not the type) of heterogeneity.

@@

We evaluate the robustness of UALE approximation, following the
evaluation framework of Section~\ref{sec:simulation-examples-2}. We
treat as ground-truth the effects computed on the full training-set
\(N=20000\) with dense fixed-size bin-splitting (\(K=80\)). Given the
big number of samples, we make the hypothesis that the approximation
with dense binning is close to the ground truth. Afterwards, we
randomly select fewer samples, \(N=1000\), and we compare UALE
approximation with all possible fixed-size approximation. We repeat
this process for \(t=30\) independent runs and we report the mean
values for \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) accross the
repetitions. In Figure~\ref{fig:ex-real-2}, we observe that UALE
achieves accurate approximations in all cases; the MAE of the
uncertainty \(\mathcal{L}_\sigma\) is (near) the best of all
fixed-size approximations.

>>>>

We define UALE and provide an interval-based formulation that
partitions the feature domain into non-overlapping intervals
(bin-splitting) and defines the bin-effect and the
bin-uncertainty. Then, we present UALE approximation, formally proving
the bin-splitting requirements for an unbiased estimation of the
uncertainty. Based on this, we present an optimization method to
determine an appropriate variable-width
partitioning. Figure~\ref{fig:UALE-figure} illustrates an UALE
visualization. The upper plot shows the ALE curve and its uncertainty
as the shaded area around the curve. The lower plot shows the bins
obtained after UALE's partitioning and the effects and the uncertainty
per bin.

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{concept_figure/fig-1.pdf}
  \caption{Illustration of UALE feature effect plot.}
  \label{fig:UALE-figure}
\end{figure}

@@

UALE defines the effect as a pair that consists of the average effect
and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
and the accumulated uncertainty, as defined in
Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates a
toy example, where \(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and the black-box function is
\(f(\xb)= \sin(2 \pi x_1) + x_1x_2\).

@@

In this simulation, we focus on illustrating the advantages of UALE's
variable-size bin-splitting against fixed-size one.

\paragraph{Example set-up.}
The purpose of this experiment is two-fold. First, we show that UALE
provides an approximation which is close to the best fixed-size
approximation, among all possible \(K\). Second, we illustrate that
UALE favors the creation of wider intervals when possible to
facilitate the interpretation of the average effect and its
uncertainty.

To measure the accuracy of the approximation, we generate a dataset
with dense sampling (\(N=10^6\)) and we use dense fixed-size bins
(\(K=10^3\)). We treat this estimation as the ground-truth
UALE. Afterwards, we generate fewer samples (\(N=500\)) and we compare
the fixed-size estimation (for several \(K\)) against UALE's.

We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) UALE's partitioning, and
with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with \(K\)
bins. The evaluation is based on the Mean Absolute Error (MAE) of the
bin effect \(\mu\) and of the uncertainty \(\sigma\) accross bins,
i.e.,

\begin{equation}
  \label{eq:eval_met_1}
  \mathcal{L}^{\mu} = \frac{1}{|\mathcal{Z}| - 1} \sum_{k \in
  \mathcal{Z}} | \mu(z_{k-1}, z_k) - \hat{\mu}(z_{k-1}, z_k) |
\end{equation}

\begin{equation}
  \label{eq:eval_met_2}
  \mathcal{L}^{\sigma} = \frac{1}{|\mathcal{Z}| -1} \sum_{k \in
    \mathcal{Z}} | \sigma(z_{k-1}, z_k) - \hat{\sigma}(z_{k-1}, z_k) |
\end{equation}

We obtain the ground truth bin-effect, \(\mu(z_{k-1}, z_k)\) and the
ground truth uncertainty, \(\sigma(z_{k-1}, z_k)\), using the average
of the dense fixed-size bins that are within in the interval
\([z_{k-1}, z_k]\). We also provide the mean residual error
\(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \) for facilitating the
interpretation of the uncertainty error.

@@

Here, the black-box function is \(f(\mathbf{x}) = a_1x_1 + x_1x_2\),
with \(5\) different-width regions, i.e., \(a_1\) equals to
\(\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the sequence
\(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). We are interested in the effect and
the uncertainty of \(x_1\) which are
\(f^{\mathcal{GT}}_{\mu}(x_1) = a_1x_1\) in the regions defined above
and \(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\), respectively.

As we observe in the top left of Figure~\ref{fig:ex-synth-2-1}, UALE's
bin-splitting separates the fine-grained bins, e.g., intervals
\([0.4, 0.45]\), \([0.45, 0.5]\), and unites (most of) constant-effect
regions into a single bin, e.g.~region \([0.5, 1]\). This explains the
fact that UALE achieves lower MAE \(\mathcal{L}^{\mu}\),
\(\mathcal{L}^{\sigma}\) compared to fixed-size bins for all \(K\).

@@

Here, the black-box function is
\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so for \(x_1\) the effect
is \(f^{\mathcal{GT}}_{\mu}(x_1) = 4 x_1^2\) and the uncertainty is
\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\). Due to the
non-linear effect, wide bins (low \(K\)) increase the mean residual
error, \(\mathcal{L}^{\rho}\), making the uncertainty approximation
biased. This can be clearly observed in in the bottom-right of
Figure~\ref{fig:ex-synth-2-2}. On the other hand, narrow bins (high
\(K\)) lead to worse approximation due to the limited number of
samples. Interestingly, UALE's bin splitting manages to compromise
these conflicting objectives and provides an accurate estimation, as we
see in the bottom-left of Figure~\ref{fig:ex-synth-2-2}.

@@

In real-world datasets, it is infeasible to obtain the ground truth FE
for evaluation. We observe in Figure~\ref{fig:ex-real-1} that UALE and
PDP-ICE compute, roughly, a similar average effect and
uncertainty. Note that when the average effect remains (almost)
constant in a wide interval, i.e., the feature effect plot is linear,
UALE creates a single bin for this interval, which facilitates the
interpretation. This can be observed, in the bottom plots of
Figure~\ref{fig:ex-real-1}, in the interval \(x_8 \in [1.5, 6]\).

@@

We evaluate the robustness of UALE approximation, following the
evaluation framework of Section~\ref{sec:simulation-examples-2}. We
treat as ground-truth the effects computed on the full training set
(\(N_{tr}=15639\)) with dense fixed-size bin-splitting
(\(K=80\)). Given the sufficient number of samples, we make the
hypothesis that the approximation with dense binning is close to the
ground truth. Afterwards, we randomly select fewer samples,
\(N=1000\), and we compare UALE approximation against fixed-size
approximation (for all \(K\)). We repeat this process for \(t=30\)
independent runs and we report the mean values of
\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\). In
Figure~\ref{fig:ex-real-2}, we observe that UALE achieves accurate
approximations in all cases; the MAE of the uncertainty
\(\mathcal{L}^\sigma\) is close to the best of the fixed-size
approximations.
